---
title: "COVID platelets UMAP"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
library(readxl)
 library(CATALYST)
library(SingleCellExperiment)
library(diffcyt)
library(HDCytoData)
library(DT)
library(ggplot2)
library(cowplot)


```
## Input data: not activated samples
### Patients metadata with 7 cases and 11 controls
```{r}
md <- read_excel("data/meta_11vs8.xlsx")
DT::datatable(data.frame(md))
```
### Markers panel
Activation markers are indicated as "state", the rest - "type". This is a very important stratification as usually lineage markers (type) are used for clustering while functional markers (state) are used for the differential expression analysis. In the furter analysis we use CD63, CD107a, PAC1 and CD62P as state markers.

```{r}
panel <- read_excel("data/panel_umap.xlsx")
datatable(data.frame(panel))
```

### .fcs files

```{r}

files <- list.files(path = "files_na", pattern = "\\.fcs$", full.names = TRUE)
#exclude cvd004b
# very important that all loaded fcs files are in the meta data and vice versa
files <- files[-c(4,20)]

fs <- read.flowSet(files, transformation = FALSE, truncate_max_range = FALSE)
fs = fs[,panel$fcs_colname]
```

Saving the data for further analysis in Python

```{r}

outDir = "data/files_na/"
for (i in 1:length(fs)){
  write.flowSet(fs[i], outDir)
}
```



Checking if panel file fits the column names in fcs files:

```{r}
all(panel$fcs_colname %in% colnames(fs))
```

Prepare metadata and check if the filenames match:

```{r}
md$condition <- factor(md$condition, levels = c("healthy", "patient"))

md$sample_id <- factor(md$sample_id, 
                       levels = md$sample_id[order(md$condition)])
```


```{r}
ids1 <- fsApply(fs, identifier)

md = subset(md,file_name %in% ids1)
```

Constracting SingleCellExperiment object: 
```{r}
sce <- prepData(fs, panel, md, features = panel$fcs_colname, cofactor = 5)
```


```{r}
set.seed(1234)
sce <- runDR(sce, "UMAP", cells =min(n_cells(sce)), features = "type")
```

Saving the UMAP dimentions 
```{r}
dr = "UMAP"
color_by = "condition"
dims = c(1,2)
xy <- reducedDim(sce, dr)[, dims]
colnames(xy) <- c("x", "y")
df <- data.frame(colData(sce), xy)
df <- df[!(is.na(df$x) | is.na(df$y)), ]
outDir = "data/umap.csv"
write.csv(df,outDir, row.names = FALSE)

```


```{r}
cs = c('#9AB8C8','#DBA794')
plotDR(sce, "UMAP", color_by = "condition", k_pal = cs) 


```

```{r, fig.width = 10, fig.height = 8}
plot_grid(plotDR(sce, "UMAP", color_by = "CD63",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD107a",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD62P",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD154",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "PAC1",facet_by = "condition"),
  ncol = 2)

```

## Activated samples


```{r}
md <- read_excel("data/meta_11vs8_activated.xlsx")
DT::datatable(data.frame(md))
```
### Markers panel
Activation markers are indicated as "state", the rest - "type". This is a very important stratification as usually lineage markers (type) are used for clustering while functional markers (state) are used for the differential expression analysis. In the furter analysis we use CD63, CD107a, PAC1 and CD62P as state markers and the rest as type markers.

```{r}
panel <- read_excel("data/panel_umap.xlsx")
datatable(data.frame(panel))
```

### .fcs files

```{r}

files <- list.files(path = "files_activated", pattern = "\\.fcs$", full.names = TRUE)
#exclude cvd004b
# very important that all loaded fcs files are in the meta data and vice versa
files <- files[-c(4,20)]

fs <- read.flowSet(files, transformation = FALSE, truncate_max_range = FALSE)
fs = fs[,panel$fcs_colname]
```

Saving the data for further analysis in Python

```{r}

outDir = "data/files_activated/"
for (i in 1:length(fs)){
  write.flowSet(fs[i], outDir)
}
```

Checking if panel file fits the column names in fcs files:

```{r}
all(panel$fcs_colname %in% colnames(fs))
```

Prepare metadata and check if the filenames match:

```{r}
md$condition <- factor(md$condition, levels = c("healthy", "patient"))

md$sample_id <- factor(md$sample_id, 
                       levels = md$sample_id[order(md$condition)])
```


```{r}
ids1 <- fsApply(fs, identifier)

md = subset(md,file_name %in% ids1)
```

Constracting SingleCellExperiment object: 
```{r}
sce <- prepData(fs, panel, md, features = panel$fcs_colname, cofactor = 5)

set.seed(1234)
sce <- runDR(sce, "UMAP", cells = min(n_cells(sce)), features = "type")

```
Saving the UMAP dimentions 
```{r}
dr = "UMAP"
color_by = "condition"
dims = c(1,2)
xy <- reducedDim(sce, dr)[, dims]
colnames(xy) <- c("x", "y")
df <- data.frame(colData(sce), xy)
df <- df[!(is.na(df$x) | is.na(df$y)), ]
outDir = "data/umap_activated.csv"
write.csv(df,outDir, row.names = FALSE)

```
```{r}
plotDR(sce, "UMAP", color_by = "condition")

```

```{r, fig.width = 10, fig.height = 8}
plot_grid(plotDR(sce, "UMAP", color_by = "CD63",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD107a",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD62P",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "CD154",facet_by = "condition"),
  plotDR(sce, "UMAP", color_by = "PAC1",facet_by = "condition"),
  ncol = 2)
```

